---
title: "XCMS processing of the untargeted metabolomics data"
author: "Mar Garcia-Aloy"
output: 
  BiocStyle::html_document:
    toc: true
    number_sections: false
    toc_float: true
---

```{r startpoint, include = FALSE}
startpoint <- Sys.time()
```

In this document I perform the XCMS processing of the untargeted metabolomics 
data.  
This file covers chromatographic peak detection, alignment, correspondence and peak filling.


# Preliminaries

## Parameters

```{r parameters}
polarity <- "NEG" # specify "POS" or "NEG"

MZML_PATH <- paste0("mzML_2019_02/", polarity, "_FS_fixed/")
dir.create("data/RData/", recursive = TRUE, showWarnings = FALSE)
```


## Libraries

```{r libraries, message = FALSE}
library(xcms)
```


# Data import

```{r data-import}
# Read the data
data_raw <- readMSData(
  paste0("data/", polarity, "_FS_fixed/",
         list.files(paste0("data/", polarity, "_FS_fixed/"))),
  mode = "onDisk")
```


# Peak detection

```{r peak-detection, message=FALSE}
cwp <- CentWaveParam(ppm = 20,
                     peakwidth = c(2, 40),
                     prefilter = c(3, 5000),
                     snthresh = 5,
                     noise = 1000,
                     mzdiff = 0.001,
                     integrate = 2)

xdata <- findChromPeaks(data_raw, param = cwp)
```


## Filter low-intensity peaks

```{r filter-intensity}
xdata <- refineChromPeaks(xdata, 
                          param = FilterIntensityParam(
                            nValues = 3, threshold = 5000))
```


## Peak post-processing

```{r peak-postproc, eval = TRUE}
mnp <- MergeNeighboringPeaksParam(
  expandRt = 2, 
  expandMz = 0.001, 
  ppm = 10,
  minProp = 0.66)

register(SerialParam())
xdata <- refineChromPeaks(xdata, param = mnp)
```


# Aligment

```{r aligment, eval = FALSE}
# Define the hook peaks
pdp <- PeakDensityParam(sampleGroups = rep(1, length(fileNames(xdata))),
                                           #xdata$type,
                        minFraction = 1,
                        binSize = 0.02,
                        bw = 3)
xdata <- groupChromPeaks(xdata, param = pdp)

# Perform the aligment
pgp <- PeakGroupsParam(span = 0.3,
                       #subset = which(xdata$class == "QC"),
                       #subsetAdjust = "average",
                       minFraction = 1)
xdata <- adjustRtime(xdata, param = pgp)
```


## Checking

The results from the alignment are shown below. 
To visualize the alignment results, the BPC for the whole data set is plotted.

```{r alignment-rtime-plot, fig.width = 12, fig.height = 10, eval = FALSE}
chr_raw <- chromatogram(xdata, aggregationFun = "max", 
                        adjustedRtime = FALSE)
chr_adj <- chromatogram(xdata, aggregationFun = "max")

par(mfrow = c(3, 1), mar = c(0, 4.3, 2, 0.1))
plot(chr_raw, #col = paste0(col_class[factor(chr_raw$class)], 40),
     peakType = "none", main = "BPC, raw")
plot(chr_adj, #col = paste0(col_class[factor(chr_raw$class)], 40),
     peakType = "none", main = "BPC, adjusted")
plotAdjustedRtime(xdata#, col = paste0(col_class[factor(xdata$class)],80)
                  )
```


# Correspondance

Correspondence aims to group chromatographic peaks of the same ion across samples.  
Usually, we re-use settings for the grouping performed already in the previous
section, but reduce the `minFraction` parameter in order to
allow definition of features that are present in fewer individuals. 

```{r correspondance}
pdp <- PeakDensityParam(rep(1, length(fileNames(xdata))),
                        #sampleGroups = xdata$type,
                        minFraction = 1,
                        binSize = 0.02,
                        bw = 2)
xdata <- groupChromPeaks(xdata, param = pdp)
```


# Peak filling

Next I fill-in missing peak data (i.e., data for samples in which no 
peak was detected).

```{r peak-filling}
fcp <- ChromPeakAreaParam()
xdata <- fillChromPeaks(xdata, param = fcp)
```


# Save data

```{r save}
save(xdata, file = paste0(
  "data/RData/data_XCMS_", polarity, ".RData"))
```


# Session information

```{r session}
Sys.time()-startpoint

devtools::session_info()
```
